---
title: "Introduction to jsonr"
author: "CLU"
date: "2025-04-13"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to jsonr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(jsonr)
library(jsonlite)
```

# Introduction

The `jsonr` package provides advanced tools for manipulating JSON data in R. While the `jsonlite` package offers excellent functionality for converting between JSON and R data structures, `jsonr` extends these capabilities with specialized functions for more complex JSON manipulation tasks:

1. **Adding elements to specific paths** in JSON objects using dot notation
2. **Inserting properties into nested array elements** with fine-grained control over positioning
3. **Utility functions** for validating, comparing, and formatting JSON data

This vignette provides an introduction to the main functionality of the `jsonr` package.

## Why jsonr?

Working with complex, nested JSON structures can be challenging, especially when:

- You need to add or modify elements at specific paths
- You need to insert properties in array elements at specific positions
- You need to process nested arrays with dynamic values

The `jsonr` package provides specialized functions to handle these tasks efficiently and intuitively.

# Basic JSON Manipulation with add_json_element()

The `add_json_element()` function allows you to add an element to a specific path in a JSON object or file. The path is specified using dot notation, making it easy to target nested structures.

## Adding an Element to a Simple JSON Object

Let's start with a simple example:

```{r}
# Create a simple JSON object
json_data <- fromJSON('{"person": {"name": "John Doe", "age": 30}}', simplifyVector = FALSE)

# Add an email element to the person object
result <- add_json_element(json_data, "person", "email", "john@example.com")

# View the result
jsonlite::toJSON(result, pretty = TRUE, auto_unbox = TRUE)
```

## Adding an Element to the Root Level

To add an element at the root level, use an empty string as the path:

```{r}
# Add a status element to the root level
result <- add_json_element(json_data, "", "status", "active")

# View the result
jsonlite::toJSON(result, pretty = TRUE, auto_unbox = TRUE)
```

## Adding an Element to a Nested Path

You can add elements to deeper paths using dot notation:

```{r}
# Add an address element with nested structure
address <- list(street = "123 Main St", city = "New York", zip = "10001")
result <- add_json_element(json_data, "person", "address", address)

# View the result
jsonlite::toJSON(result, pretty = TRUE, auto_unbox = TRUE)
```

## Creating Missing Path Components

By default, the function will throw an error if a path component doesn't exist. However, you can set `create_path = TRUE` to create missing components:

```{r}
# Try to add a phone number to a non-existent contact structure
result <- add_json_element(json_data, "person.contact.phone", "mobile", "+1234567890", create_path = TRUE)

# View the result
jsonlite::toJSON(result, pretty = TRUE, auto_unbox = TRUE)
```

If you don't set `create_path = TRUE` and the path doesn't exist, you'll get an error:

```{r, error=TRUE}
# This will fail because the path doesn't exist
add_json_element(json_data, "person.contact.phone", "mobile", "+1234567890")
```

## Controlling Element Position

You can control the position of the new element within its parent structure:

```{r}
# Add an element at the beginning (position = 1)
result <- add_json_element(json_data, "person", "id", 123, position = 1)

# View the result (note that "id" is now the first element in "person")
jsonlite::toJSON(result, pretty = TRUE, auto_unbox = TRUE)
```

## Working with JSON Files

The `add_json_element()` function can also work directly with JSON files:

```{r, eval=FALSE}
# Add an element to a JSON file
add_json_element("data.json", "settings", "language", "en")

# Specify a different output file to avoid overwriting the original
add_json_element("data.json", "settings", "theme", "dark", output_file_path = "output.json")
```

# Manipulating Nested Arrays with insert_json_property()

The `insert_json_property()` function is designed for more complex scenarios where you need to insert properties into elements of nested arrays.

## Basic Usage

Let's start with a simple example:

```{r}
# Complex JSON with nested arrays
json_str <- '{
  "items": [
    {
      "id": 1,
      "name": "Laptop",
      "tests": [
        {"name": "Performance", "score": 85},
        {"name": "Battery", "score": 90}
      ]
    },
    {
      "id": 2,
      "name": "Smartphone",
      "tests": [
        {"name": "Performance", "score": 80},
        {"name": "Battery", "score": 75}
      ]
    }
  ]
}'

# Parse the JSON
complex_data <- fromJSON(json_str, simplifyVector = FALSE)

# Insert a "passed" property into all test elements
result <- insert_json_property(
  complex_data,
  array_path = c("items", "tests"),
  property_name = "passed",
  property_value = TRUE,
  position_type = "last"  # Use "last" as a safe default
)

# View the result
jsonlite::toJSON(result, pretty = TRUE, auto_unbox = TRUE)
```

## Controlling Property Position

You can control the position of the new property within each element:

```{r}
# Insert a property at the beginning of each element
result <- insert_json_property(
  complex_data,
  array_path = c("items", "tests"),
  property_name = "testId",
  property_value = 1,
  position_type = "first"
)

# View the result
jsonlite::toJSON(result, pretty = TRUE, auto_unbox = TRUE)

# Insert a property after a specific property
result <- insert_json_property(
  complex_data,
  array_path = c("items", "tests"),
  property_name = "testId",
  property_value = 1,
  position_type = "after",
  position_ref = "name"
)

# View the result
jsonlite::toJSON(result, pretty = TRUE, auto_unbox = TRUE)
```

## Dynamic Property Values

One of the most powerful features of `insert_json_property()` is the ability to use a function to dynamically generate property values based on context:

```{r}
# Insert a property with a dynamic value based on the parent element
result <- insert_json_property(
  complex_data,
  array_path = c("items", "tests"),
  property_name = "itemName",
  property_value = function(elem, index, parent, parent_index, root) {
    return(parent$name)  # Use the name of the parent item
  },
  position_type = "last"  # Use "last" as the position_type
)

# View the result
jsonlite::toJSON(result, pretty = TRUE, auto_unbox = TRUE)
```

## Filtering Elements

You can also use a filter function to selectively apply the property to certain elements:

```{r}
# Insert a property only for test elements with a score >= 80
result <- insert_json_property(
  complex_data,
  array_path = c("items", "tests"),
  property_name = "excellent",
  property_value = TRUE,
  position_type = "last",  # Use "last" as the position_type
  filter_fn = function(elem) {
    return(elem$score >= 80)
  }
)

# View the result
jsonlite::toJSON(result, pretty = TRUE, auto_unbox = TRUE)
```

# Conclusion

The `jsonr` package provides powerful tools for manipulating JSON data in R, complementing the existing functionality in packages like `jsonlite`. 

Key features include:

- **Path-based element addition** with `add_json_element()`
- **Nested array manipulation** with `insert_json_property()`
- **Fine-grained control** over element and property positioning
- **Dynamic property values** based on context
- **Selective modification** with filtering

These capabilities make `jsonr` a valuable tool for working with complex JSON structures in R.

For more advanced usage examples, see the "Advanced JSON Manipulation" vignette.